---
title: "Exploratory Visualization of Fraud Risk Patterns in Bank Transaction Data"
author: "Ngoc Khanh Vy Huynh, Keyao Li"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
fontsize: 11pt
geometry: margin=0.85in
mainfont: "Arial"
monofont: "Menlo"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{needspace}
  - \usepackage{caption}
  - \usepackage{sectsty}
  - \allsectionsfont{\sffamily}
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(kableExtra)
  library(patchwork)
})

# Locate the project whether knitting starts in the project root or paper/.
project_root <- if (file.exists("data/bank_fraud.csv")) {
  normalizePath(".")
} else {
  normalizePath("..")
}

data_dir <- file.path(project_root, "data")
figure_dir <- file.path(project_root, "figures")
paper_dir <- file.path(project_root, "paper")
summary_dir <- file.path(paper_dir, "summaries")
raw_data_path <- file.path(data_dir, "bank_fraud.csv")

dir.create(figure_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paper_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(summary_dir, showWarnings = FALSE, recursive = TRUE)

theme_set(
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", colour = "#17365D"),
      plot.subtitle = element_text(colour = "#52606D"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.8,
  fig.pos = "H",
  dpi = 300
)

fraud_colours <- c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")

save_plot <- function(plot, filename, width = 7.5, height = 5) {
  ggsave(
    filename = file.path(figure_dir, filename), plot = plot,
    width = width, height = height, dpi = 300, bg = "white"
  )
  plot
}

save_table <- function(table, filename) {
  write_csv(table, file.path(summary_dir, filename))
  invisible(table)
}

style_table <- function(table, caption, digits = NULL, align = NULL) {
  kableExtra::kbl(
    table, format = "latex", booktabs = TRUE, caption = caption,
    digits = digits, align = align
  ) %>%
    kableExtra::kable_styling(
      full_width = FALSE,
      font_size = 9,
      latex_options = c("hold_position", "striped", "scale_down")
    )
}

# Standard summary used for categorical fraud-rate comparisons.
fraud_rate_summary <- function(data, group_variable) {
  data %>%
    group_by({{ group_variable }}) %>%
    summarise(
      total_transactions = n(),
      fraud_count = sum(is_fraud == 1, na.rm = TRUE),
      fraud_rate = mean(is_fraud == 1, na.rm = TRUE),
      .groups = "drop"
    )
}
```

# 1. Data Preparation

The full dataset was used for descriptive summaries and aggregated fraud-rate
calculations. A small stratified sample was created only for code submission
and sample-based plots. The dataset contains 1,000,000 transactions and 26
variables; `is_fraud` identifies fraudulent (1) and non-fraudulent (0)
transactions.

```{r read-data}
bank_fraud <- read_csv(raw_data_path, show_col_types = FALSE)

required_variables <- c(
  "transaction_id", "is_fraud", "fraud_type", "is_international",
  "is_night_transaction", "is_weekend", "pin_changed_recently",
  "customer_age", "transaction_amount", "account_balance", "credit_score",
  "hour_of_day", "failed_attempts"
)

missing_required <- setdiff(required_variables, names(bank_fraud))
if (length(missing_required) > 0) {
  stop("Missing required variables: ", paste(missing_required, collapse = ", "))
}

dataset_dimensions <- tibble(
  Measure = c("Number of rows", "Number of columns"),
  Value = c(comma(nrow(bank_fraud)), as.character(ncol(bank_fraud)))
)

variable_groups <- tribble(
  ~`Variable group`, ~`Variables`, ~`Role in analysis`,
  "Identifiers", "transaction_id, customer_id",
    "Used for duplicate checks, not visualization",
  "Date and time", "transaction_date, transaction_time, hour_of_day",
    "Used for temporal transaction patterns",
  "Numerical variables",
    "transaction_amount, account_balance, customer_age, credit_score, distance_from_home_km, time_since_last_txn_hrs, failed_attempts",
    "Used for distribution and fraud-status comparisons",
  "Binary indicators",
    "is_weekend, is_night_transaction, is_international, pin_changed_recently, is_fraud",
    "Converted into readable labels for fraud-rate visualizations",
  "Categorical variables",
    "country, city, merchant_category, payment_method, device_type, fraud_type",
    "Used for category-level comparison and fraud-rate analysis"
)

style_table(dataset_dimensions, "Dataset Dimensions", align = c("l", "r"))
```

```{r data-quality, results='hide'}
missing_values_summary <- tibble(
  variable = names(bank_fraud),
  missing_count = map_int(bank_fraud, ~ sum(is.na(.x))),
  missing_percentage = map_dbl(
    bank_fraud, ~ 100 * sum(is.na(.x)) / length(.x)
  )
)

duplicate_transaction_id_count <- sum(duplicated(bank_fraud$transaction_id))
duplicate_full_row_count <- sum(duplicated(bank_fraud))

data_quality_summary <- tribble(
  ~quality_check, ~issue_count,
  "Missing values (all variables)", sum(missing_values_summary$missing_count),
  "Duplicate transaction_id values", duplicate_transaction_id_count,
  "Duplicate full rows", duplicate_full_row_count,
  "customer_age < 18", sum(bank_fraud$customer_age < 18, na.rm = TRUE),
  "transaction_amount <= 0", sum(bank_fraud$transaction_amount <= 0, na.rm = TRUE),
  "account_balance < 0", sum(bank_fraud$account_balance < 0, na.rm = TRUE),
  "credit_score outside 300 to 850",
    sum(bank_fraud$credit_score < 300 | bank_fraud$credit_score > 850, na.rm = TRUE),
  "hour_of_day outside 0 to 23",
    sum(bank_fraud$hour_of_day < 0 | bank_fraud$hour_of_day > 23, na.rm = TRUE),
  "failed_attempts < 0", sum(bank_fraud$failed_attempts < 0, na.rm = TRUE)
) %>%
  mutate(issue_percentage = 100 * issue_count / nrow(bank_fraud))

save_table(missing_values_summary, "missing_values_summary.csv")
save_table(data_quality_summary, "data_quality_summary.csv")
```

Data-quality checks found no missing values, duplicate transaction IDs,
duplicate full rows, or invalid values in the selected key variables, so no row
removal or imputation was required before visualization.

```{r preparation-summaries}
fraud_summary <- bank_fraud %>%
  summarise(
    total_transactions = n(),
    fraud_count = sum(is_fraud == 1, na.rm = TRUE),
    non_fraud_count = sum(is_fraud == 0, na.rm = TRUE),
    fraud_rate = mean(is_fraud == 1, na.rm = TRUE)
  )

fraud_type_summary <- bank_fraud %>%
  mutate(fraud_type = replace_na(fraud_type, "Missing")) %>%
  count(fraud_type, name = "count") %>%
  mutate(percentage = 100 * count / sum(count)) %>%
  arrange(desc(count))

key_numeric_variables <- c(
  "customer_age", "credit_score", "account_age_years", "account_balance",
  "transaction_amount", "num_prev_transactions",
  "transaction_freq_monthly", "distance_from_home_km",
  "time_since_last_txn_hrs", "failed_attempts"
)

numeric_summary <- bank_fraud %>%
  select(all_of(key_numeric_variables)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    minimum = min(value, na.rm = TRUE),
    first_quartile = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    mean = mean(value, na.rm = TRUE),
    third_quartile = quantile(value, 0.75, na.rm = TRUE),
    maximum = max(value, na.rm = TRUE),
    missing_count = sum(is.na(value)),
    .groups = "drop"
  )

save_table(fraud_summary, "fraud_summary.csv")
save_table(fraud_type_summary, "fraud_type_summary.csv")
save_table(numeric_summary, "numeric_summary.csv")

fraud_summary_display <- fraud_summary %>%
  transmute(
    `Total transactions` = comma(total_transactions),
    `Fraud count` = comma(fraud_count),
    `Non-fraud count` = comma(non_fraud_count),
    `Fraud rate` = percent(fraud_rate, accuracy = 0.01)
  )

numeric_summary_display <- numeric_summary %>%
  transmute(
    Variable = str_replace_all(variable, "_", " ") %>% str_to_title(),
    Minimum = comma(minimum, accuracy = 0.01),
    Median = comma(median, accuracy = 0.01),
    Mean = comma(mean, accuracy = 0.01),
    Maximum = comma(maximum, accuracy = 0.01)
  )

transaction_numeric_display <- numeric_summary_display %>%
  filter(Variable %in% str_to_title(str_replace_all(c(
    "transaction_amount", "account_balance", "distance_from_home_km",
    "time_since_last_txn_hrs", "failed_attempts"
  ), "_", " ")))

customer_numeric_display <- numeric_summary_display %>%
  filter(Variable %in% str_to_title(str_replace_all(c(
    "customer_age", "credit_score", "account_age_years",
    "num_prev_transactions", "transaction_freq_monthly"
  ), "_", " ")))

style_table(fraud_summary_display, "Overall Fraud Summary",
            align = c("r", "r", "r", "r"))
```

There are 55,255 fraudulent and 944,745 non-fraudulent transactions, giving an
overall fraud rate of 5.53%. Because this outcome is highly imbalanced, later
comparisons use within-group fraud rates rather than raw fraud counts. Readable
labels were created for fraud status, international status, night status,
weekend status, and recent PIN change.

```{r prepare-data, include=FALSE}
bank_fraud_prepared <- bank_fraud %>%
  mutate(
    fraud_status = factor(
      if_else(is_fraud == 1, "Fraud", "Non-fraud"),
      levels = c("Non-fraud", "Fraud")
    ),
    transaction_type = factor(
      if_else(is_international == 1, "International", "Domestic"),
      levels = c("Domestic", "International")
    ),
    night_status = factor(
      if_else(is_night_transaction == 1, "Night", "Day"),
      levels = c("Day", "Night")
    ),
    weekend_status = factor(
      if_else(is_weekend == 1, "Weekend", "Weekday"),
      levels = c("Weekday", "Weekend")
    ),
    pin_change_status = factor(
      if_else(pin_changed_recently == 1,
              "Recent PIN Change", "No Recent PIN Change"),
      levels = c("No Recent PIN Change", "Recent PIN Change")
    ),
    across(
      c(country, city, merchant_category, payment_method, device_type, fraud_type),
      as.factor
    )
  )

write_csv(bank_fraud_prepared,
          file.path(data_dir, "bank_fraud_prepared.csv"))

set.seed(123)
sample_bank_fraud <- bank_fraud_prepared %>%
  group_by(is_fraud) %>%
  slice_sample(n = 1000) %>%
  ungroup()

write_csv(sample_bank_fraud,
          file.path(data_dir, "sample_bank_fraud.csv"))
```

The prepared dataset retains all observations. A reproducible sample of 1,000
fraudulent and 1,000 non-fraudulent transactions was selected with seed 123 for
compact code submission; full-dataset results are reported below.

# 2. Dataset Overview

```{r overview-distributions, fig.width=7, fig.height=3.2, out.width='82%', fig.cap="Compact overview of (A) outcome imbalance and (B) fraud-type composition."}
overview_fraud_counts <- bank_fraud_prepared %>%
  count(fraud_status)

p_overview_fraud <- ggplot(
  overview_fraud_counts,
  aes(fraud_status, n, fill = fraud_status)
) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = comma(n)), vjust = -0.35, size = 3.8) +
  scale_fill_manual(values = fraud_colours) +
  scale_y_continuous(labels = label_number(big.mark = ","),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Transaction status", y = "Transactions")

overview_fraud_types <- bank_fraud_prepared %>%
  filter(!is.na(fraud_type), fraud_type != "None") %>%
  count(fraud_type, sort = TRUE) %>%
  mutate(fraud_type = fct_reorder(fraud_type, n))

p_overview_types <- ggplot(overview_fraud_types, aes(fraud_type, n)) +
  geom_col(fill = "#E45756", width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(x = "Fraud type", y = "Fraudulent transactions")

overview_combined <- p_overview_fraud + p_overview_types +
  plot_layout(widths = c(0.85, 1.3)) +
  plot_annotation(tag_levels = "A")

ggsave(file.path(figure_dir, "overview_fraud_distribution.png"),
       p_overview_fraud, width = 5.5, height = 3.2, dpi = 300, bg = "white")
ggsave(file.path(figure_dir, "overview_fraud_type_distribution.png"),
       p_overview_types, width = 6, height = 3.2, dpi = 300, bg = "white")
save_plot(overview_combined, "overview_combined.png", 7, 3.2)
```

Panel A shows that fraud is a minority outcome, representing 5.53% of all
transactions; this imbalance motivates the use of rates in later comparisons.
Panel B shows that the six fraud types are comparatively even, with no single
type dominating the fraudulent subset. Together, the panels provide compact
background context without competing visually with the risk-pattern results.

# 3. Univariate Analysis

## Numerical variables

The following plots describe one variable at a time and do not compare fraud
groups. Transaction amount uses a logarithmic x-axis because of its strong
right skew.

```{r univariate-numeric-plots, fig.show='hold', out.width='48%', fig.width=5.5, fig.height=3.2}
p_amount <- ggplot(bank_fraud_prepared, aes(transaction_amount)) +
  geom_histogram(bins = 50, fill = "#4C78A8", colour = "white") +
  scale_x_log10(labels = label_dollar()) +
  labs(title = "Transaction Amount", x = "Transaction amount (log scale)",
       y = "Number of transactions")

p_balance <- ggplot(bank_fraud_prepared, aes(account_balance)) +
  geom_histogram(bins = 50, fill = "#59A14F", colour = "white") +
  scale_x_log10(labels = label_dollar()) +
  labs(title = "Account Balance", x = "Account balance (log scale)",
       y = "Number of transactions")

p_age <- ggplot(bank_fraud_prepared, aes(customer_age)) +
  geom_histogram(binwidth = 2, fill = "#F28E2B", colour = "white") +
  labs(title = "Customer Age", x = "Age (years)", y = "Number of transactions")

p_distance <- ggplot(bank_fraud_prepared, aes(distance_from_home_km)) +
  geom_histogram(bins = 50, fill = "#B279A2", colour = "white") +
  labs(title = "Distance from Home", x = "Distance (km)",
       y = "Number of transactions")

p_time_since <- ggplot(bank_fraud_prepared, aes(time_since_last_txn_hrs)) +
  geom_histogram(bins = 50, fill = "#76B7B2", colour = "white") +
  labs(title = "Time Since Last Transaction", x = "Hours",
       y = "Number of transactions")

p_hour <- bank_fraud_prepared %>%
  count(hour_of_day) %>%
  ggplot(aes(hour_of_day, n)) +
  geom_col(fill = "#4C78A8") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Transactions by Hour", x = "Hour of day",
       y = "Number of transactions")

p_failed <- bank_fraud_prepared %>%
  count(failed_attempts) %>%
  ggplot(aes(factor(failed_attempts), n)) +
  geom_col(fill = "#E15759") +
  scale_y_continuous(labels = comma) +
  labs(title = "Failed Attempts", x = "Number of failed attempts",
       y = "Number of transactions")

numeric_plots <- list(
  transaction_amount = p_amount, account_balance = p_balance,
  customer_age = p_age, distance_from_home_km = p_distance,
  time_since_last_txn_hrs = p_time_since, hour_of_day = p_hour,
  failed_attempts = p_failed
)

iwalk(numeric_plots, ~ ggsave(
  file.path(figure_dir, paste0("univariate_", .y, ".png")),
  .x, width = 7.5, height = 5, dpi = 300, bg = "white"
))

walk(numeric_plots[c(
  "transaction_amount", "failed_attempts"
)], print)
```

Transaction amount is strongly right-skewed, which supports logarithmic scaling
when comparing fraud groups. Failed attempts are concentrated at zero and one,
with higher counts comparatively uncommon; this provides context for the later
fraud-rate result. Univariate plots describe dataset composition but do not
directly answer which characteristics are associated with fraud. Therefore, the
bivariate and multivariate sections provide the main evidence for identifying
fraud-risk patterns.

```{r univariate-categorical-plots, include=FALSE}
ordered_bar <- function(data, variable, title, x_label, fill = "#4C78A8",
                        exclude_none = FALSE) {
  plot_data <- data %>%
    filter(!is.na({{ variable }}))
  if (exclude_none) {
    plot_data <- plot_data %>% filter(as.character({{ variable }}) != "None")
  }
  plot_data %>%
    count({{ variable }}, sort = TRUE) %>%
    mutate(category = fct_reorder(as.factor({{ variable }}), n)) %>%
    ggplot(aes(category, n)) +
    geom_col(fill = fill, width = 0.72) +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(title = title, x = x_label, y = "Number of transactions")
}

categorical_plots <- list(
  fraud_type = ordered_bar(bank_fraud_prepared, fraud_type,
    "Fraud Types", "Fraud type", "#E45756", TRUE),
  merchant_category = ordered_bar(bank_fraud_prepared, merchant_category,
    "Merchant Categories", "Merchant category", "#59A14F"),
  payment_method = ordered_bar(bank_fraud_prepared, payment_method,
    "Payment Methods", "Payment method", "#4C78A8"),
  device_type = ordered_bar(bank_fraud_prepared, device_type,
    "Device Types", "Device type", "#B279A2"),
  country = ordered_bar(bank_fraud_prepared, country,
    "Transactions by Country", "Country", "#F28E2B")
)

iwalk(categorical_plots, ~ ggsave(
  file.path(figure_dir, paste0("univariate_", .y, ".png")),
  .x, width = 8, height = 5.5, dpi = 300, bg = "white"
))

```

# 4. Bivariate Fraud Analysis

All comparisons below report the total number of transactions, fraud count,
and fraud rate. The interpretation uses associative language and does not imply
that any transaction characteristic causes fraud.

## Binary transaction characteristics

```{r bivariate-binary, fig.width=7, fig.height=5.4, out.width='90%', fig.cap="Fraud rates across four binary transaction characteristics."}
binary_specs <- list(
  night_status = list(var = "night_status", title = "Time of Day",
                      x = "Time-of-day status"),
  transaction_type = list(var = "transaction_type", title = "Transaction Type",
                          x = "Transaction type"),
  pin_change_status = list(var = "pin_change_status", title = "PIN Change Status",
                           x = "PIN-change status"),
  weekend_status = list(var = "weekend_status", title = "Weekend Status",
                        x = "Calendar status")
)

binary_results <- imap(binary_specs, function(spec, file_stub) {
  summary_table <- bank_fraud_prepared %>%
    group_by(.data[[spec$var]]) %>%
    summarise(
      total_transactions = n(), fraud_count = sum(is_fraud == 1),
      fraud_rate = mean(is_fraud == 1), .groups = "drop"
    )
  names(summary_table)[1] <- spec$var
  save_table(summary_table, paste0("bivariate_", file_stub, "_summary.csv"))

  x_labels <- if (file_stub == "pin_change_status") {
    c("No Recent PIN Change" = "No PIN Change",
      "Recent PIN Change" = "PIN Changed")
  } else {
    waiver()
  }

  plot <- ggplot(summary_table,
                 aes(x = .data[[spec$var]], y = fraud_rate,
                     fill = .data[[spec$var]])) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(aes(label = percent(fraud_rate, accuracy = 0.01)),
              vjust = -0.35, size = 3.4) +
    scale_x_discrete(labels = x_labels) +
    scale_y_continuous(labels = label_percent(accuracy = 0.1),
                       expand = expansion(mult = c(0, 0.12))) +
    labs(title = spec$title, x = NULL, y = "Fraud rate") +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      axis.text.x = element_text(size = 8, angle = 12, hjust = 1)
    )

  ggsave(file.path(figure_dir, paste0("bivariate_", file_stub, ".png")),
         plot, width = 7.2, height = 5, dpi = 300, bg = "white")
  plot
})

binary_combined <- wrap_plots(binary_results, ncol = 2) +
  plot_annotation(tag_levels = "A")
save_plot(binary_combined, "bivariate_binary_characteristics.png", 7, 5.4)
```

International transactions, night transactions, and transactions following a
recent PIN change show visibly higher fraud rates than their comparison groups,
while weekend and weekday rates are nearly identical. The first three patterns
appear to be stronger visual indicators, whereas weekend status offers little
visual separation. These associations directly address the research question
but do not imply causation.

## Merchant category

```{r bivariate-merchant-category, fig.width=6.5, fig.height=4.2, out.width='82%', fig.cap="Within-category fraud rates and transaction counts by merchant context."}
merchant_fraud <- fraud_rate_summary(bank_fraud_prepared, merchant_category) %>%
  arrange(desc(fraud_rate))
save_table(merchant_fraud, "bivariate_merchant_category_summary.csv")

p_merchant_fraud <- merchant_fraud %>%
  mutate(merchant_category = fct_reorder(merchant_category, fraud_rate)) %>%
  ggplot(aes(merchant_category, fraud_rate)) +
  geom_col(fill = "#59A14F") +
  geom_text(aes(label = percent(fraud_rate, accuracy = 0.1)),
            hjust = -0.05, size = 2.8) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.14))) +
  labs(x = "Merchant category", y = "Fraud rate")

save_plot(p_merchant_fraud, "bivariate_merchant_category.png", 8.5, 6)
```

ATM withdrawal, jewelry, and crypto exchange transactions form a distinct upper
group near 9%, while most other merchant categories cluster closer to the
overall rate. Each merchant category contains approximately 66,000
transactions, so the rate differences are not driven by very small groups.
Merchant context therefore appears strongly associated with fraud risk and is
one of the stronger candidate figures.

```{r bivariate-payment-method, include=FALSE}
payment_fraud <- fraud_rate_summary(bank_fraud_prepared, payment_method) %>%
  arrange(desc(fraud_rate))
save_table(payment_fraud, "bivariate_payment_method_summary.csv")

p_payment_fraud <- payment_fraud %>%
  mutate(payment_method = fct_reorder(payment_method, fraud_rate)) %>%
  ggplot(aes(payment_method, fraud_rate)) +
  geom_col(fill = "#4C78A8") +
  coord_flip() +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(title = "Fraud Rate by Payment Method", x = "Payment method",
       y = "Fraud rate")

save_plot(p_payment_fraud, "bivariate_payment_method.png", 8, 5.5)
```

```{r bivariate-device-type, include=FALSE}
device_fraud <- fraud_rate_summary(bank_fraud_prepared, device_type) %>%
  arrange(desc(fraud_rate))
save_table(device_fraud, "bivariate_device_type_summary.csv")

p_device_fraud <- device_fraud %>%
  mutate(device_type = fct_reorder(device_type, fraud_rate)) %>%
  ggplot(aes(device_type, fraud_rate)) +
  geom_col(fill = "#B279A2") +
  coord_flip() +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(title = "Fraud Rate by Device Type", x = "Device type",
       y = "Fraud rate")

save_plot(p_device_fraud, "bivariate_device_type.png", 8, 5)
```

\Needspace{0.48\textheight}

## Transaction amount by fraud status

```{r bivariate-amount, fig.width=5.5, fig.height=3.2, out.width='68%', fig.cap="Transaction amount by fraud status on a logarithmic scale."}
amount_fraud_summary <- bank_fraud_prepared %>%
  group_by(fraud_status) %>%
  summarise(
    total_transactions = n(), fraud_count = sum(is_fraud == 1),
    fraud_rate = mean(is_fraud == 1),
    median_transaction_amount = median(transaction_amount, na.rm = TRUE),
    mean_transaction_amount = mean(transaction_amount, na.rm = TRUE),
    .groups = "drop"
  )
save_table(amount_fraud_summary, "bivariate_transaction_amount_summary.csv")

p_amount_fraud <- ggplot(
  bank_fraud_prepared,
  aes(fraud_status, transaction_amount, fill = fraud_status)
) +
  geom_boxplot(width = 0.62, outlier.alpha = 0.04, show.legend = FALSE) +
  scale_fill_manual(values = fraud_colours) +
  scale_y_log10(labels = label_dollar()) +
  labs(x = "Transaction status", y = "Transaction amount (log scale)")

save_plot(p_amount_fraud, "bivariate_transaction_amount.png")
```

The logarithmic boxplot shows substantial overlap between fraudulent and
non-fraudulent transaction amounts. Transaction amount alone therefore provides
weak visual separation between the two statuses. This cautionary result shows
that an intuitively relevant variable need not be a strong standalone indicator.

\Needspace{0.48\textheight}

## Failed attempts

```{r bivariate-failed-attempts, fig.width=6, fig.height=3.4, out.width='75%', fig.cap="Fraud rate by number of failed attempts."}
failed_fraud <- fraud_rate_summary(bank_fraud_prepared, failed_attempts) %>%
  arrange(failed_attempts)
save_table(failed_fraud, "bivariate_failed_attempts_summary.csv")

p_failed_fraud <- ggplot(failed_fraud, aes(failed_attempts, fraud_rate)) +
  geom_line(linewidth = 1, colour = "#E45756") +
  geom_point(size = 2.5, colour = "#E45756") +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(x = "Number of failed attempts", y = "Fraud rate")

save_plot(p_failed_fraud, "bivariate_failed_attempts.png")
```

Fraud rates remain relatively low at zero or one failed attempt and rise sharply
when the count reaches two or more. The ordered separation is stronger than in
most other comparisons. Repeated failed attempts therefore appear strongly
associated with elevated fraud risk, making this pattern useful for the figure
shortlist.

\Needspace{0.48\textheight}

## Hour of day

```{r bivariate-hour, fig.width=6, fig.height=3.4, out.width='75%', fig.cap="Hourly fraud rate across the 24-hour transaction cycle."}
hour_fraud <- fraud_rate_summary(bank_fraud_prepared, hour_of_day) %>%
  arrange(hour_of_day)
save_table(hour_fraud, "bivariate_hour_of_day_summary.csv")

p_hour_fraud <- ggplot(hour_fraud, aes(hour_of_day, fraud_rate)) +
  geom_line(linewidth = 1, colour = "#4C78A8") +
  geom_point(size = 1.8, colour = "#4C78A8") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(x = "Hour of day", y = "Fraud rate")

save_plot(p_hour_fraud, "bivariate_hour_of_day.png", 8, 5)
```

Fraud rates appear higher during late-night and early-morning hours and remain
lower through most daytime periods. This agrees with the night-status comparison
and shows a strong, repeated timing pattern across adjacent hours. Transaction
timing is therefore one of the stronger candidate patterns, although the result
is associational rather than causal.

```{r bivariate-distance, include=FALSE}
distance_fraud_summary <- bank_fraud_prepared %>%
  group_by(fraud_status) %>%
  summarise(
    total_transactions = n(), fraud_count = sum(is_fraud == 1),
    fraud_rate = mean(is_fraud == 1),
    median_distance_km = median(distance_from_home_km, na.rm = TRUE),
    mean_distance_km = mean(distance_from_home_km, na.rm = TRUE),
    .groups = "drop"
  )
save_table(distance_fraud_summary, "bivariate_distance_from_home_summary.csv")

p_distance_fraud <- ggplot(
  bank_fraud_prepared,
  aes(fraud_status, distance_from_home_km, fill = fraud_status)
) +
  geom_boxplot(width = 0.62, outlier.alpha = 0.04, show.legend = FALSE) +
  scale_fill_manual(values = fraud_colours) +
  scale_y_log10(labels = label_number(suffix = " km")) +
  labs(title = "Distance from Home by Fraud Status",
       x = "Transaction status", y = "Distance from home (log scale)")

save_plot(p_distance_fraud, "bivariate_distance_from_home.png")
```

\clearpage

# 5. Multivariate Fraud Pattern Analysis

```{r multivariate-patterns, fig.width=7, fig.height=5.8, out.width='86%', fig.cap="Multivariate fraud patterns: (A) hourly rates by transaction type and (B) merchant-category rates across hours."}
hour_transaction_fraud <- bank_fraud_prepared %>%
  group_by(hour_of_day, transaction_type) %>%
  summarise(
    total_transactions = n(),
    fraud_count = sum(is_fraud == 1),
    fraud_rate = mean(is_fraud == 1),
    .groups = "drop"
  )
save_table(hour_transaction_fraud,
           "multivariate_hour_transaction_type_summary.csv")

p_hour_transaction <- ggplot(
  hour_transaction_fraud,
  aes(hour_of_day, fraud_rate, colour = transaction_type)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.4) +
  scale_colour_manual(values = c("Domestic" = "#4C78A8",
                                 "International" = "#E45756")) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(title = "Hourly Pattern by Transaction Type", x = "Hour of day",
       y = "Fraud rate", colour = NULL)

merchant_hour_fraud <- bank_fraud_prepared %>%
  group_by(merchant_category, hour_of_day) %>%
  summarise(
    total_transactions = n(),
    fraud_count = sum(is_fraud == 1),
    fraud_rate = mean(is_fraud == 1),
    .groups = "drop"
  )
save_table(merchant_hour_fraud,
           "multivariate_merchant_hour_summary.csv")

p_merchant_hour <- ggplot(
  merchant_hour_fraud,
  aes(hour_of_day, fct_rev(merchant_category), fill = fraud_rate)
) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_x_continuous(breaks = seq(0, 23, 3), expand = c(0, 0)) +
  scale_fill_viridis_c(labels = label_percent(accuracy = 0.1),
                       option = "C", direction = -1) +
  labs(title = "Merchant Context Across the Day", x = "Hour of day",
       y = "Merchant category", fill = "Fraud rate") +
  theme(axis.text.y = element_text(size = 7.5))

multivariate_combined <- p_hour_transaction / p_merchant_hour +
  plot_layout(heights = c(0.9, 1.25)) +
  plot_annotation(tag_levels = "A")

ggsave(file.path(figure_dir, "multivariate_hour_transaction_type.png"),
       p_hour_transaction, width = 7, height = 3.3, dpi = 300, bg = "white")
ggsave(file.path(figure_dir, "multivariate_merchant_hour_heatmap.png"),
       p_merchant_hour, width = 7, height = 4.2, dpi = 300, bg = "white")
save_plot(multivariate_combined, "multivariate_combined.png", 7, 5.8)
```

Panel A shows that international transactions have higher fraud rates than
domestic transactions across most hours. Both groups also show higher fraud
rates during late-night and early-morning periods. Panel B adds merchant context
and shows that ATM Withdrawal, Jewelry, and Crypto Exchange remain visually
higher across multiple hours. Together, these plots suggest that transaction
timing and merchant context jointly describe higher fraud-risk patterns in this
synthetic dataset.

\Needspace{0.30\textheight}

# 6. Key Findings and Visual Profile

Fraud is a minority class with an overall rate of 5.53%. The clearest visual
patterns involve failed attempts, transaction timing, international status,
recent PIN changes, and merchant category. Transactions with two or more failed
attempts show substantially higher fraud rates, while late-night and
early-morning periods remain elevated across domestic and international groups.
ATM withdrawal, jewelry, and crypto exchange transactions appear as the
higher-risk merchant contexts. In contrast, transaction amount, payment method,
device type, and distance from home show weaker visual separation and are less
visually distinctive than the stronger bivariate results.

# 7. Notes for Figure Selection

```{r figure-selection, echo=FALSE}
figure_selection <- tribble(
  ~`Figure name`, ~`Analysis type`, ~Recommendation,
  "Binary transaction characteristics", "Bivariate", "Must use",
  "Fraud rate by merchant category", "Bivariate", "Must use",
  "Fraud rate by failed attempts", "Bivariate", "Must use",
  "Fraud rate by hour of day", "Bivariate", "Must use",
  "Merchant-by-hour heatmap", "Multivariate", "Must use",
  "Fraud/non-fraud distribution", "Dataset overview", "Maybe -- small overview only",
  "Transaction amount by fraud status", "Bivariate", "Maybe -- cautionary weak finding"
)

style_table(
  figure_selection,
  "Figure Shortlist for Team Discussion",
  align = c("l", "l", "l")
)
```

The shortlist prioritizes figures that add distinct evidence rather than
repeating the same pattern. The combined overview and transaction-amount result
remain optional because they provide useful context rather than strong
separation.

**Limitation.** Because the dataset is synthetic, the results should be
interpreted as visual patterns within this dataset rather than direct evidence
about real-world banking behaviour. Location variables are treated as
categorical features, not as verified geographic relationships.
