---
title: "Exploratory Visualization of Fraud Risk Patterns in Bank Transaction Data"
affiliation:
  wide: true
  author_multiline:
    - line:
        - name: "Ngoc Khanh Vy Huynh, Student Number: 0908036"
    - line:
        - name: "Keyao Li, Student Number: 0895585"
abstract: |
  Bank transaction fraud is difficult to examine because fraudulent observations are rare and risk patterns may differ across transaction contexts. This paper uses exploratory data analysis and visualization to investigate which transaction characteristics show visible associations with fraud risk in a synthetic dataset of one million bank transactions. The analysis combines descriptive summaries, within-group fraud-rate comparisons, distribution plots, and a merchant-by-hour heatmap. Fraud accounts for 55,255 transactions, or 5.53% of the full dataset, creating a strongly imbalanced outcome. The clearest visual patterns involve repeated failed attempts, late-night and early-morning timing, international transaction status, recent PIN changes, and merchant category. Transactions with two or more failed attempts have substantially higher fraud rates, while ATM withdrawal, jewelry, and crypto exchange transactions appear elevated across multiple hours. In contrast, weekend status and account balance provide weaker or more cautionary visual separation; location variables were inspected but not retained as main figures. Visualization therefore identifies candidate fraud-risk indicators before predictive modeling, although all findings are associative and based on synthetic data.
keywords: "bank transaction fraud, exploratory data analysis, data visualization, class imbalance, fraud rate"
output:
  rticles::ieee_article:
    keep_tex: true
    with_dblfloatfix: true
    citation_package: natbib
bibliography: support/references.bib
biblio-style: IEEEtran
natbiboptions: "numbers"
link-citations: true
header-includes:
  - \usepackage[section]{placeins}
  - \usepackage{booktabs}
---

<!--
TODO BEFORE FINAL SUBMISSION
1. Add the verified Kaggle dataset title, creator, URL, license, and access date.
2. Confirm author affiliations and email addresses required by the course.
3. Verify figure legibility and the final page count in the IEEE template.
-->

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(scales)
  library(patchwork)
  library(janitor)
  library(knitr)
  library(kableExtra)
})

source("support/helper_functions.R")

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

theme_set(paper_theme())
fraud_colours <- c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")
```

```{r read-and-prepare-data, include=FALSE}
data_choice <- choose_transaction_data(
  full_path = "../../bank_fraud.csv",
  sample_path = "support/sample_transactions.csv"
)
transactions_raw <- readr::read_csv(data_choice$path, show_col_types = FALSE)
df <- prepare_transactions(transactions_raw)
using_sample_only <- data_choice$using_sample

sample_path <- "support/sample_transactions.csv"
if (file.exists(sample_path)) {
  sample_df <- readr::read_csv(sample_path, show_col_types = FALSE) |>
    prepare_transactions()
} else {
  set.seed(123)
  sample_df <- df |>
    dplyr::filter(is_fraud %in% c(0L, 1L)) |>
    dplyr::group_by(is_fraud) |>
    dplyr::slice_sample(n = min(1000L, dplyr::n())) |>
    dplyr::ungroup()
}

dataset_summary <- df |>
  dplyr::summarise(
    transactions = dplyr::n(),
    variables = ncol(transactions_raw),
    fraud = sum(is_fraud == 1L, na.rm = TRUE),
    non_fraud = sum(is_fraud == 0L, na.rm = TRUE),
    fraud_rate = mean(is_fraud == 1L, na.rm = TRUE)
  )

overall_fraud_rate <- dataset_summary$fraud_rate

missing_count <- sum(is.na(df))
duplicate_transaction_ids <- if ("transaction_id" %in% names(df)) {
  sum(duplicated(df$transaction_id))
} else {
  NA_integer_
}
duplicate_rows <- sum(duplicated(df))

invalid_values <- tibble::tibble(
  check = c(
    "customer_age < 18",
    "transaction_amount <= 0",
    "account_balance < 0",
    "credit_score outside 300--850",
    "hour_of_day outside 0--23",
    "failed_attempts < 0"
  ),
  count = c(
    if ("customer_age" %in% names(df)) sum(df$customer_age < 18, na.rm = TRUE) else NA,
    if ("transaction_amount" %in% names(df)) sum(df$transaction_amount <= 0, na.rm = TRUE) else NA,
    sum(df$account_balance < 0, na.rm = TRUE),
    if ("credit_score" %in% names(df)) {
      sum(df$credit_score < 300 | df$credit_score > 850, na.rm = TRUE)
    } else NA,
    sum(df$hour_of_day < 0 | df$hour_of_day > 23, na.rm = TRUE),
    sum(df$failed_attempts < 0, na.rm = TRUE)
  )
)
```

```{r sample-warning, results="asis"}
if (using_sample_only) {
  cat(
    "> **Demonstration-data warning:** `data/transactions.csv` was not found. ",
    "This render uses the balanced sample only. Its figures demonstrate the ",
    "code workflow and must not be interpreted as population prevalence.\n\n"
  )
}
```

# Introduction

Transaction fraud monitoring is important because suspicious activity must be
identified within a large volume of legitimate banking transactions. A central
challenge is class imbalance: fraudulent transactions usually form a small
minority, so raw counts can obscure meaningful differences between transaction
groups. A second challenge is interpretability. Before developing a predictive
model, analysts need to understand whether observable transaction
characteristics display stable and communicable patterns.

Exploratory data analysis (EDA) supports this early stage by summarizing data
quality, revealing skewed distributions, and comparing fraud rates across
behavioural and contextual variables. Visual analysis does not establish
causation, but it can identify candidate indicators for later statistical
testing or predictive modeling. It is also useful for communication: analysts,
risk managers, and other stakeholders can inspect whether an apparent pattern
is consistent across groups or driven by class size, skewness, or a small
subset of observations.

This paper asks: *Which transaction characteristics show visible associations
with fraud risk in bank transaction data?* Its contribution is a focused visual
comparison across authentication behaviour, transaction timing, international
status, recent PIN changes, merchant context, and selected numerical variables.
Because fraud is uncommon, the analysis emphasizes within-group rates and
proportions rather than raw fraud counts. This paper follows the project
proposal's exploratory scope by prioritizing exploratory visualization over
predictive modeling. Although the proposal mentioned that a simple classifier
could be developed if time permitted, the primary objective was to identify and
explain visible fraud-risk patterns through EDA. The analysis therefore
establishes a reproducible evidence base for describing candidate variables and
interactions without making causal or predictive claims.

The remainder of the paper reviews relevant fraud-detection and imbalanced-data
research, describes the dataset and reproducible EDA methodology, reports
univariate, bivariate, and multivariate visual findings, and concludes with
practical implications, limitations, and directions for validation.

# Previous Work

Fraud-detection research has examined transaction behaviour, authentication
signals, temporal activity, and contextual variables as possible inputs to
monitoring systems [@abdallah2016]. Anomaly-detection research considers how
rare observations can be distinguished from dominant normal behaviour
[@chandola2009]. Research on imbalanced learning also shows why overall accuracy
and raw counts are inadequate when the positive class is uncommon [@he2009].

The imbalance problem is especially important in financial transaction data.
When legitimate transactions greatly outnumber fraudulent ones, a method can
appear accurate while failing to identify the minority class. Recent fraud
detection research therefore emphasizes resampling, minority-class synthesis,
ensemble methods, and precision-recall-oriented evaluation
[@fiore2019; @kennedy2024; @bekereci2024]. These studies are primarily
predictive, but their motivation also applies to EDA: population counts and
balanced-sample proportions answer different questions and must not be
interchanged.

Fraud observations may also overlap substantially with legitimate behaviour.
Li *et al.* describe the combination of class imbalance and class overlap as a
central challenge because fraudulent transactions are often designed to
resemble ordinary activity [@li2021]. This supports examining multiple
dimensions rather than expecting one variable to separate the classes.
Authentication behaviour, transaction context, timing, and merchant type may
jointly carry more information than an isolated financial amount.

EDA complements predictive methods by exposing distributions, group
differences, and possible interactions before a model is fitted
[@tukey1977]. Bar charts can compare categorical rates, distribution plots can
show overlap, and heatmaps can expose patterns across contextual dimensions.
However, many fraud studies move quickly to classifier comparison. The present
study instead emphasizes a transparent visual workflow that distinguishes
population-rate figures from balanced-sample figures and reports both strong
and weak patterns. This study remains exploratory: it identifies visible
candidate patterns but does not estimate causal effects or claim predictive
performance.

# Methodology

## Dataset and Variables

The full synthetic dataset was obtained from the *Bank Transaction Fraud
Detection Dataset* hosted on Kaggle [@kaggleBankFraud]. It contains 1,000,000
transactions and 26 variables.
The binary outcome, `is_fraud`, equals one for a fraudulent transaction and zero
otherwise. The full dataset contains 55,255 fraudulent and 944,745
non-fraudulent transactions, corresponding to an overall fraud rate of 5.53%.

```{r table-dataset-summary, results="asis"}
summary_display <- dataset_summary |>
  dplyr::transmute(
    Transactions = scales::comma(transactions),
    Variables = variables,
    Fraud = scales::comma(fraud),
    `Non-fraud` = scales::comma(non_fraud),
    `Fraud rate` = scales::percent(fraud_rate, accuracy = 0.01)
  )

knitr::kable(
  summary_display,
  format = "latex",
  booktabs = TRUE,
  caption = if (using_sample_only) {
    "Active demonstration sample summary"
  } else {
    "Full transaction dataset summary"
  }
) |>
  kableExtra::kable_styling(
    full_width = FALSE,
    latex_options = c("hold_position", "scale_down")
  )
```

Variables examined include transaction type, hour of day, night and weekend
indicators, merchant category, failed attempts, account balance, distance from
home, city, and recent PIN change. Full-data summaries and fraud-rate
calculations are used whenever `data/transactions.csv` is available.

The variables can be organized into four conceptual groups. Transaction context
includes domestic or international status, merchant category, payment method,
device type, and location. Temporal context includes hour, night status, and
weekend status. Authentication and account context includes failed attempts,
recent PIN change, account balance, and credit-related fields. The binary
outcome `is_fraud` provides the common basis for comparing these groups.

## Data Preparation

Column names were standardized, the binary outcome was converted to integer
form, and readable factors were created for Fraud/Non-fraud,
International/Domestic, Night/Day, Weekend/Weekday, and PIN Changed/No PIN
Change. Data-quality checks covered missingness, repeated identifiers,
duplicate complete rows, and invalid ranges for selected key variables. In the
full dataset, these checks found no missing values, duplicate transaction IDs,
duplicate rows, or invalid values, so no removal or imputation was required.

A balanced sample containing 1,000 fraud and 1,000 non-fraud observations is
used for selected distributional and sample-composition figures. Fraud rates
by failed attempts and transaction type are calculated from the full dataset.
Because the balanced sample is deliberately constructed, it cannot estimate
population prevalence.

## Reproducible Analysis Workflow

The paper is written as an executable R Markdown document. The document reads
the CSV data, standardizes variables, calculates every displayed summary, and
generates each figure during rendering. Supporting functions are stored in a
small sourced R file. No static screenshot is used as an analytical result.
This design makes the relationship among data, calculations, figures, and
written interpretations auditable.

The full dataset is used for counts and fraud-rate estimates, including the
failed-attempt comparison in Figure 5. The small balanced sample is used only
for selected distributional or sample-composition figures and where a runnable
submission sample is required. Every sample-based caption explicitly warns
that its shares do not estimate population prevalence. This separation
prevents the 50% fraud share of the balanced sample from being confused with
the 5.53% fraud rate in the full data.

## Analytical Measures

For each category, the descriptive fraud rate is the number of fraudulent
transactions divided by the total number of transactions in that category.
Within-group sample proportions are calculated separately for fraud and
non-fraud cases so that the majority class does not dominate a distribution
plot.

## Visualization Strategy

Bar charts compare rates across binary and categorical variables. A faceted
point-range plot compares full-data fraud rates across discrete failed-attempt
values and transaction types, with Wilson confidence intervals used to
communicate uncertainty. A heatmap combines hour of day and merchant category.
For grouped full-data comparisons, total transactions, fraud counts, and fraud
rates are calculated directly from `df`. Findings are described as visual
associations, not causal effects.

# Exploratory Data Analysis Design

The EDA proceeded in three stages. First, univariate checks established dataset
size, class balance, variable ranges, missingness, and duplicate status. This
stage determined whether cleaning or imputation was necessary and revealed the
extent of outcome imbalance. Second, bivariate comparisons evaluated how fraud
rates or within-group distributions changed across transaction type, time of
day, recent PIN change, weekend status, failed attempts, account balance, and
city. Third, a multivariate heatmap combined merchant category with hour of day
to determine whether two contextual dimensions revealed patterns hidden by
separate marginal summaries.

The visual encodings were selected for the measurement scale and analytical
question. Counts are appropriate for showing class imbalance; rate bars support
fair categorical comparisons; a mosaic plot displays both transaction-type
frequency and within-type composition; faceted point-and-line displays show
fraud rates across discrete failed-attempt values while preserving transaction
context and uncertainty; violin and box plots show density, median, and
overlap; and a heatmap displays a two-dimensional rate surface. This staged
design supports interpretation without presenting EDA as proof of causality or
predictive accuracy.

# Results

## Outcome Imbalance and Dataset Context

```{r fig-class-imbalance, fig.pos="!htbp", fig.width=3.45, fig.height=2.55, out.width="\\columnwidth", fig.cap="Class distribution of transaction status. Fraud represents 5.53% of the full dataset, indicating a highly imbalanced outcome."}
class_counts <- df |>
  dplyr::count(fraud_label, name = "transactions")

p_class <- ggplot(class_counts, aes(fraud_label, transactions, fill = fraud_label)) +
  geom_col(width = 0.64, show.legend = FALSE) +
  geom_text(
    aes(label = scales::comma(transactions)),
    vjust = -0.35,
    size = 3
  ) +
  scale_fill_manual(values = fraud_colours) +
  scale_y_continuous(
    labels = scales::label_number(big.mark = ","),
    expand = expansion(mult = c(0, 0.10))
  ) +
  labs(x = "Transaction status", y = "Transactions")

p_class
```

Fraud is a minority outcome: the full-data rate is 5.53%, and the non-fraud
class is more than seventeen times larger. Consequently, later comparisons use
within-group fraud rates rather than raw counts. The data-quality assessment
found no missing values, duplicate transaction identifiers, duplicate complete
rows, or out-of-range values among the checked fields. Therefore, the observed
patterns are not artifacts of an imputation or row-removal procedure.

## Transaction Type and Sample Composition

```{r fig-transaction-type-mosaic, fig.pos="!htbp", fig.width=3.45, fig.height=2.75, out.width="\\columnwidth", fig.cap="Transaction type and fraud-group membership in the balanced sample. Rectangle widths reflect sample counts; sample shares do not estimate population prevalence."}
transaction_mosaic <- sample_df |>
  dplyr::filter(!is.na(transaction_type), !is.na(fraud_label)) |>
  dplyr::count(transaction_type, fraud_label, name = "n") |>
  dplyr::group_by(transaction_type) |>
  dplyr::mutate(
    type_total = sum(n),
    share = n / type_total,
    ymax = cumsum(share),
    ymin = dplyr::lag(ymax, default = 0)
  ) |>
  dplyr::ungroup()

type_widths <- transaction_mosaic |>
  dplyr::distinct(transaction_type, type_total) |>
  dplyr::arrange(transaction_type) |>
  dplyr::mutate(
    xmax = cumsum(type_total) / sum(type_total),
    xmin = dplyr::lag(xmax, default = 0),
    xmid = (xmin + xmax) / 2
  )

transaction_mosaic <- transaction_mosaic |>
  dplyr::left_join(type_widths, by = c("transaction_type", "type_total")) |>
  dplyr::mutate(
    label = paste0(
      fraud_label, "\n",
      "n=", n, "\n",
      scales::percent(share, accuracy = 0.1)
    )
  )

ggplot(transaction_mosaic) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fraud_label),
    colour = "white", linewidth = 0.45
  ) +
  geom_text(
    aes(x = xmid, y = (ymin + ymax) / 2, label = label),
    colour = "white", size = 2.4, fontface = "bold"
  ) +
  scale_fill_manual(values = fraud_colours, guide = "none") +
  scale_x_continuous(
    breaks = type_widths$xmid,
    labels = paste0(type_widths$transaction_type, "\n(n=", type_widths$type_total, ")"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  labs(x = "Transaction type", y = "Within-type sample composition") +
  theme(panel.grid = element_blank(), axis.text.x = element_text(size = 7))
```

Within the balanced sample, international transactions contain a larger fraud
share than domestic transactions. Because fraud and non-fraud cases were sampled
equally, this figure describes sample composition and is not a population fraud
rate. Specifically, fraud represents 68.1% of sampled international
transactions but 45.1% of sampled domestic transactions. The difference is
directionally consistent with the full-data international fraud-rate result,
but the percentages in Fig. 2 must remain sample-composition measures.

## Binary Transaction Characteristics

```{r fig-binary-characteristics, fig.pos="!htbp", fig.width=3.45, fig.height=4.35, out.width="\\columnwidth", fig.cap="Fraud rates across selected binary transaction characteristics. Night transactions and recent PIN changes show higher fraud rates, while weekend status shows little separation."}
binary_plot <- function(data, variable, title) {
  rate_data <- fraud_rate_by(data, {{ variable }})

  ggplot(rate_data, aes({{ variable }}, fraud_rate, fill = {{ variable }})) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(
      aes(label = scales::percent(fraud_rate, accuracy = 0.01)),
      vjust = -0.30,
      size = 2.7
    ) +
    scale_y_continuous(
      labels = scales::label_percent(accuracy = 0.1),
      limits = c(0, max(rate_data$fraud_rate) * 1.24),
      expand = expansion(mult = c(0, 0))
    ) +
    scale_fill_manual(values = c("#4C78A8", "#E45756")) +
    labs(title = title, x = NULL, y = "Fraud rate") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 9),
      axis.text.x = element_text(size = 7.5, angle = 10, hjust = 1)
    )
}

binary_panels <- list(
  binary_plot(df, time_of_day, "Time of Day"),
  binary_plot(df, pin_change_status, "PIN Change Status"),
  binary_plot(df, weekend_status, "Weekend Status")
)

patchwork::wrap_plots(binary_panels, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```

Night transactions and transactions following a recent PIN change show visibly
higher fraud rates than their comparison groups. Night transactions have a
7.96% fraud rate compared with 4.07% during the day. Transactions following a
recent PIN change reach 8.36%, compared with 5.28% when no recent change is
recorded. Weekend and weekday rates are nearly equal at 5.57% and 5.51%,
respectively. This contrast is useful because it shows that not every intuitive
context variable separates fraud. These differences are candidate visual
indicators, not evidence of causation.

## Account Balance

The fraud group has a somewhat higher median balance in the balanced sample,
but the distributions overlap substantially. Account balance therefore provides
weaker standalone separation than failed attempts or transaction context. The
sample median is approximately \$8,157 for fraud and \$7,390 for non-fraud, but
the interquartile ranges and violin densities occupy much of the same region.
This is a useful negative result: a shift in central tendency does not imply
that the variable is an effective standalone discriminator.

```{r fig-account-balance, fig.pos="!htbp", fig.width=3.45, fig.height=2.75, out.width="0.95\\columnwidth", fig.cap="Account-balance distribution shapes by fraud group in the balanced sample. Violin width shows density and inset boxes show the median and interquartile range; the balanced sample does not estimate population prevalence."}
balance_summary <- sample_df |>
  dplyr::filter(!is.na(account_balance), !is.na(fraud_label)) |>
  dplyr::group_by(fraud_label) |>
  dplyr::summarise(
    n = dplyr::n(),
    median_balance = stats::median(account_balance),
    .groups = "drop"
  )

ggplot(sample_df, aes(fraud_label, log1p(account_balance), fill = fraud_label)) +
  geom_violin(trim = FALSE, alpha = 0.58, show.legend = FALSE) +
  geom_boxplot(
    width = 0.13, outlier.shape = NA, fill = "white",
    colour = "#333333", linewidth = 0.35
  ) +
  geom_text(
    data = balance_summary,
    aes(
      x = fraud_label,
      y = min(log1p(sample_df$account_balance), na.rm = TRUE),
      label = paste0("n=", n, "\nmedian=$", scales::comma(round(median_balance)))
    ),
    inherit.aes = FALSE, vjust = 1.15, size = 2.2
  ) +
  scale_fill_manual(values = fraud_colours) +
  labs(x = "Sample group", y = "Log account balance")
```

## Failed Attempts Across Transaction Contexts

```{r fig-failed-attempts, fig.env="figure*", fig.pos="!t", fig.width=7.1, fig.height=2.8, out.width="0.94\\textwidth", fig.cap="Full-data fraud rates by failed attempts and transaction type. Points show descriptive fraud rates, and vertical bars show 95% Wilson confidence intervals. The largest increase occurs between one and two failed attempts in both transaction contexts."}
# Wilson intervals are calculated directly for each full-data context cell.
wilson_z <- stats::qnorm(0.975)

failed_context_rates <- df |>
  dplyr::filter(
    !is.na(failed_attempts),
    !is.na(transaction_type),
    !is.na(is_fraud)
  ) |>
  dplyr::group_by(transaction_type, failed_attempts) |>
  dplyr::summarise(
    total_transactions = dplyr::n(),
    fraud_count = sum(is_fraud == 1L),
    fraud_rate = fraud_count / total_transactions,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    wilson_denominator = 1 + wilson_z^2 / total_transactions,
    wilson_centre = (fraud_rate + wilson_z^2 / (2 * total_transactions)) /
      wilson_denominator,
    wilson_half_width = wilson_z * sqrt(
      fraud_rate * (1 - fraud_rate) / total_transactions +
        wilson_z^2 / (4 * total_transactions^2)
    ) / wilson_denominator,
    lower = pmax(0, wilson_centre - wilson_half_width),
    upper = pmin(1, wilson_centre + wilson_half_width)
  ) |>
  dplyr::group_by(transaction_type) |>
  dplyr::arrange(failed_attempts, .by_group = TRUE) |>
  dplyr::mutate(
    label_height = upper + rep(c(0.006, 0.012), length.out = dplyr::n()),
    count_label = paste0("n=", scales::comma(total_transactions))
  ) |>
  dplyr::ungroup()

context_colours <- c("Domestic" = "#4C78A8", "International" = "#E45756")

ggplot(
  failed_context_rates,
  aes(failed_attempts, fraud_rate, colour = transaction_type, group = transaction_type)
) +
  geom_vline(xintercept = 1.5, linetype = "dashed", colour = "#9CA3AF", linewidth = 0.35) +
  geom_line(linewidth = 0.65) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.12, linewidth = 0.45) +
  geom_point(size = 2.0) +
  geom_text(
    aes(y = label_height, label = count_label),
    colour = "#5E6E86", size = 2.15, show.legend = FALSE
  ) +
  facet_wrap(~ transaction_type, nrow = 1) +
  scale_colour_manual(values = context_colours, guide = "none") +
  scale_x_continuous(breaks = sort(unique(failed_context_rates$failed_attempts))) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, max(failed_context_rates$label_height) * 1.08),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(x = "Failed attempts", y = "Full-data fraud rate") +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

The full-data results show a clear increase at two failed attempts. 
Domestic fraud rates rise from about 4% at zero or one attempt to roughly 14% at two or more attempts, 
while international rates increase from about 9% to around 19–20%. 
Because both curves follow a similar shape, the figure suggests a consistent failed-attempt pattern together with a higher international fraud baseline, rather than a strong interaction.

## Merchant and Time-Based Patterns

```{r fig-multivariate-patterns, fig.env="figure*", fig.pos="!t", fig.width=7.1, fig.height=5.8, out.width="0.94\\textwidth", fig.cap="Multivariate fraud patterns: (A) hourly fraud rates for domestic and international transactions and (B) merchant-category fraud rates across hours."}
hour_transaction_rates <- df |>
  dplyr::filter(
    !is.na(hour_of_day),
    !is.na(transaction_type),
    !is.na(is_fraud)
  ) |>
  dplyr::group_by(hour_of_day, transaction_type) |>
  dplyr::summarise(
    total_transactions = dplyr::n(),
    fraud_count = sum(is_fraud == 1L),
    fraud_rate = mean(is_fraud == 1L),
    .groups = "drop"
  )

p_hour_transaction <- ggplot(
  hour_transaction_rates,
  aes(hour_of_day, fraud_rate, colour = transaction_type)
) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 1.25) +
  scale_colour_manual(
    values = c("Domestic" = "#4C78A8", "International" = "#E45756"),
    name = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(x = "Hour of day", y = "Fraud rate") +
  theme(legend.position = "top")

merchant_hour_rates <- df |>
  dplyr::filter(
    !is.na(hour_of_day),
    !is.na(merchant_category),
    !is.na(is_fraud)
  ) |>
  dplyr::group_by(merchant_category, hour_of_day) |>
  dplyr::summarise(
    total_transactions = dplyr::n(),
    fraud_count = sum(is_fraud == 1L),
    fraud_rate = mean(is_fraud == 1L),
    .groups = "drop"
  )

p_merchant_hour <- ggplot(
  merchant_hour_rates,
  aes(hour_of_day, forcats::fct_rev(merchant_category), fill = fraud_rate)
) +
  geom_tile(colour = "white", linewidth = 0.12) +
  scale_x_continuous(breaks = seq(0, 23, 3), expand = c(0, 0)) +
  scale_fill_gradient(
    low = "#17324D",
    high = "#56B1F7",
    labels = scales::label_percent(accuracy = 0.1)
  ) +
  labs(x = "Hour of day", y = "Merchant category", fill = "Fraud rate") +
  theme(axis.text.y = element_text(size = 7))

p_hour_transaction / p_merchant_hour +
  patchwork::plot_layout(heights = c(0.9, 1.25)) +
  patchwork::plot_annotation(tag_levels = "A")
```

Panel A shows that international transactions have higher fraud rates than
domestic transactions across most hours within this synthetic dataset. Both
series also rise during late-night and early-morning periods. Panel B adds
merchant context: ATM withdrawal, jewelry, and crypto exchange remain visually
elevated across several hours, whereas most other categories cluster at lower
rates. Aggregated across hours, the three highlighted categories have fraud
rates of approximately 8.65--8.74%, compared with roughly 4.62--4.84% for most
other merchant categories. Together, the panels show an exploratory pattern in
which transaction timing, international status, and merchant context are
visually associated with different fraud rates. These associations are
candidate indicators within the synthetic dataset, not causal evidence.

## Summary of Visual Findings

The strongest visible associations involve repeated failed attempts,
late-night and early-morning timing, international status, recent PIN changes,
and merchant category. Account balance, weekend status, payment method, device
type, and distance from home show weaker separation or require cautious
interpretation.

Additional variables mentioned in the proposal, including payment method,
device type, transaction amount, distance from home, and other location-related
fields, were inspected during EDA; however, they showed weaker or less central
visual separation and are therefore summarized rather than displayed as main
figures within the IEEE page limit.

\FloatBarrier

# Discussion

## Interpretation and Practical Relevance

EDA identifies candidate features, reveals the effect of class imbalance, and
shows which variables have limited standalone separation. Stronger patterns
could inform feature selection, monitoring rules, interactions, or later
predictive experiments. Visual prominence alone does not establish statistical
significance or operational usefulness.

The results suggest a hierarchy of candidate information. Authentication
behaviour is represented by failed attempts and recent PIN change; transaction
context is represented by international status and merchant category; and
temporal context is represented by night or early-morning activity. These
signals are interpretable and can be reviewed by a human analyst. In contrast,
weekend status and account balance provide weaker separation. Including weak
results is important because an EDA should narrow the candidate set rather than
merely collect visually interesting charts.

For an operational fraud-monitoring workflow, the findings could motivate
rules or model features such as an indicator for two or more failed attempts,
an interaction between merchant category and hour, or additional review when a
recent PIN change coincides with an unusual context. Such uses would require
validation on out-of-sample real transactions, threshold selection based on
false-positive costs, and continuous monitoring for changing fraud behaviour.
The figures should therefore be viewed as hypothesis-generating tools, not as
decision rules.

## Limitations

The dataset is synthetic, so the findings describe patterns within this dataset
rather than verified real-world banking behaviour. Location variables are
treated as categorical features, not validated geographic relationships. EDA
cannot establish causation, and deliberately balanced sample plots cannot
estimate population prevalence. The associations should be validated with
statistical analysis and predictive evaluation that accounts for class
imbalance.

The analysis is also cross-sectional. It does not examine how relationships
change across days, customers, or evolving fraud strategies. Aggregated rates
may hide repeated activity by the same account. Finally, no classifier,
train-test split, or cost-sensitive evaluation is included because the proposal
identified modeling only as an optional extension if time permitted, while the
primary deliverable remained EDA and visualization.

# Conclusions

This paper used code-generated exploratory visualizations to examine which bank
transaction characteristics show visible associations with fraud risk. Fraud
represents 5.53% of the full dataset, making rate-based comparisons essential.
Repeated failed attempts, night and early-morning timing, international status,
recent PIN changes, and merchant category provide the clearest visual patterns.
Weekend status and account balance are weaker or more cautionary findings.

The study demonstrates why visual design and sampling language matter in an
imbalanced problem. Full-data rates describe prevalence, while balanced-sample
figures describe group composition and distributional differences. Maintaining
this distinction prevents overstated conclusions and makes the analysis
reproducible.

Future work should validate these exploratory patterns using independent or
real transaction data and examine whether the visible relationships remain
stable over time. Predictive modeling may be considered as a separate extension
after the exploratory findings are validated; it is not part of the present
paper's contribution.

# References
