---
title: "Exploratory Data Analysis: Customer and Segment Patterns"
author: "Vy"
date: "`r Sys.Date()`"
format: pdf
execute:
  warning: false
  message: false
---
```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)})

input_data_path <- "data/sample_bank_fraud.csv"
figures_dir <- "figures"
tables_dir <- "tables"

dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(tables_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
# Load packages needed for the future EDA report.
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

<!-- Placeholder: State the purpose, scope, and complementary focus of this EDA. -->

# Dataset Overview

<!-- Placeholder: Describe the sample dataset, relevant variables, and analysis scope. -->

# Customer Profile Analysis

Customer age is explored as a foundational customer-profile characteristic.
This descriptive comparison examines whether age distributions differ between
fraud and non-fraud transactions in the balanced analysis sample; it does not
imply that customer age causes fraud.

```{r customer-age-summary}

customer_age_data <- read_csv(
  input_data_path,
  col_select = c(customer_age, is_fraud),
  show_col_types = FALSE
) %>%
  mutate(
    fraud_status = factor(
      if_else(is_fraud == 1, "Fraud", "Non-fraud"),
      levels = c("Non-fraud", "Fraud")
    )
  )

customer_age_summary <- customer_age_data %>%
  group_by(fraud_status) %>%
  summarise(
    sample_size = n(),
    mean_age = mean(customer_age, na.rm = TRUE),
    median_age = median(customer_age, na.rm = TRUE),
    standard_deviation = sd(customer_age, na.rm = TRUE),
    minimum_age = min(customer_age, na.rm = TRUE),
    maximum_age = max(customer_age, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  customer_age_summary,
  digits = 2,
  caption = "Customer-age summary by fraud status"
)
```

The following faceted histogram provides a visual comparison of customer-age
distributions between the two fraud-status groups.

```{r customer-age-histogram, fig.width=8, fig.height=4.5, fig.cap="Distribution of customer age by fraud status in the balanced analysis sample."}
# Standard faceted-histogram style for future continuous-variable comparisons.
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

customer_age_plot <- ggplot(
  customer_age_data,
  aes(x = customer_age, fill = fraud_status)
) +
  geom_histogram(binwidth = 5, colour = "white", alpha = 0.85) +
  facet_wrap(~ fraud_status, nrow = 1) +
  labs(
    x = "Customer age (years)",
    y = "Number of transactions",
    fill = "Transaction status"
  )

ggsave(
  filename = file.path(figures_dir, "customer_age_distribution.png"),
  plot = customer_age_plot,
  width = 9,
  height = 5.5,
  dpi = 300,
  bg = "white"
)

customer_age_plot
```

The figure compares customer-age distributions between fraud and non-fraud
groups in a sample balanced by design. It is intended only for descriptive
comparison and should not be used to draw causal conclusions or estimate
population fraud prevalence.

# Customer Behavior Analysis

<!-- Placeholder: Explore transaction-frequency and account-behavior variables. -->

# Geographic Analysis

<!-- Placeholder: Explore country and city as categorical transaction segments. -->

# Multivariate Segment Analysis

<!-- Placeholder: Examine future relationships across customer, behavior, and geographic segments. -->

# Key Findings

<!-- Placeholder: Summarize evidence-based findings after the EDA is completed. -->

# Business Insights

<!-- Placeholder: Translate future findings into relevant business considerations. -->

# Conclusion

<!-- Placeholder: Recap the analysis, its limitations, and possible next steps. -->
