---
title: "Sample Verification Report: Exploratory Fraud Risk Patterns"
author:
  - "Ngoc Khanh Vy Huynh"
  - "Keyao Li"
format:
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Purpose

This report demonstrates that the submitted source code can read, prepare,
summarize, and visualize the supplied sample data.

> **Important:** The supplied file is a deliberately balanced sample containing
> 1,000 fraud and 1,000 non-fraud observations. Results in this report verify
> the code workflow only. They do not estimate population prevalence and are
> not expected to reproduce the full-data percentages in the final paper.

# Setup and data preparation

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
library(janitor)
library(knitr)

source("support/helper_functions.R")

sample_raw <- readr::read_csv(
  "data/sample_transactions.csv",
  show_col_types = FALSE
)
sample_df <- prepare_transactions(sample_raw)
```

# Input checks

```{r}
required_columns <- c(
  "is_fraud", "hour_of_day", "failed_attempts",
  "merchant_category", "account_balance",
  "is_international", "is_night_transaction"
)

stopifnot(
  nrow(sample_df) == 2000L,
  all(required_columns %in% names(sample_df)),
  sum(sample_df$is_fraud == 1L) == 1000L,
  sum(sample_df$is_fraud == 0L) == 1000L
)

sample_summary <- sample_df |>
  summarise(
    Rows = n(),
    Variables = ncol(sample_raw),
    Fraud = sum(is_fraud == 1L),
    `Non-fraud` = sum(is_fraud == 0L),
    `Missing outcome values` = sum(is.na(is_fraud))
  )

knitr::kable(sample_summary)
```

# Class composition

```{r}
class_summary <- sample_df |>
  count(fraud_label, name = "Transactions") |>
  mutate(Share = Transactions / sum(Transactions))

ggplot(class_summary, aes(fraud_label, Transactions, fill = fraud_label)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = paste0(comma(Transactions), " (", percent(Share), ")")),
    vjust = -0.4
  ) +
  scale_fill_manual(values = c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Class Composition of the Supplied Sample",
    x = NULL,
    y = "Transactions"
  ) +
  paper_theme(base_size = 11)
```

# Transaction-type composition

```{r}
transaction_type_summary <- sample_df |>
  filter(!is.na(transaction_type)) |>
  count(transaction_type, fraud_label, name = "Transactions") |>
  group_by(transaction_type) |>
  mutate(WithinTypeShare = Transactions / sum(Transactions)) |>
  ungroup()

knitr::kable(transaction_type_summary)

ggplot(
  transaction_type_summary,
  aes(transaction_type, WithinTypeShare, fill = fraud_label)
) +
  geom_col() +
  scale_fill_manual(values = c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Fraud-Group Composition Within Transaction Type",
    subtitle = "Balanced-sample description; not a population fraud rate",
    x = "Transaction type",
    y = "Within-type sample share",
    fill = "Sample group"
  ) +
  paper_theme(base_size = 11)
```

# Failed-attempt distribution

```{r}
failed_attempt_summary <- sample_df |>
  filter(!is.na(failed_attempts)) |>
  count(fraud_label, failed_attempts, name = "Transactions") |>
  group_by(fraud_label) |>
  mutate(WithinGroupShare = Transactions / sum(Transactions)) |>
  ungroup()

ggplot(
  failed_attempt_summary,
  aes(factor(failed_attempts), WithinGroupShare, fill = fraud_label)
) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Failed Attempts by Sample Group",
    x = "Failed attempts",
    y = "Within-group sample share",
    fill = "Sample group"
  ) +
  paper_theme(base_size = 11)
```

# Account-balance distribution

```{r}
ggplot(sample_df, aes(fraud_label, log1p(account_balance), fill = fraud_label)) +
  geom_violin(alpha = 0.65, trim = FALSE, show.legend = FALSE) +
  geom_boxplot(width = 0.16, outlier.shape = NA, show.legend = FALSE) +
  scale_fill_manual(values = c("Non-fraud" = "#4C78A8", "Fraud" = "#E45756")) +
  labs(
    title = "Account-Balance Distribution in the Supplied Sample",
    x = "Sample group",
    y = "Log account balance"
  ) +
  paper_theme(base_size = 11)
```

# Verification result

The supplied sample was read successfully, all required columns were present,
the balanced class counts were confirmed, and the submitted helper functions
produced the tables and visualizations in this report.

**Sample code verification passed.**
